Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-18 paper/main.tex
edits through p 13

\author{Andrew C. Heusser\textsuperscript{1, 2, \textdagger}, Paxton C. Fitzpatrick\textsuperscript{1, \textdagger}, and Jeremy R. Manning\textsuperscript{1, *}\\\textsuperscript{1}Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\\textsuperscript{2}Akili Interactive\\Boston, MA 02110\\\textsuperscript{\textdagger}Denotes equal contribution\\\textsuperscript{*}Corresponding author: jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract}
The ways our experiences unfold over time may be characterized by unique \textit{trajectories} through geometric representational spaces.  In word embedding spaces, for example, each coordinate reflects a concept and nearby coordinates (in Euclidean distance) reflect semantically related concepts.  We propose a framework for projecting naturalistic experiences into word embedding spaces, such that the conceptual content of each moment of an experience, and how different moments of the experience relate, are reflected by the \textit{shape} of the experience's trajectory.  By projecting memories of those experiences into the same spaces, one may then geometrically compare the shape of the original experience's trajectory to the shape of how it is remembered later.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the original experience's trajectory.  This translates qualitative questions about how we remember naturalistic experiences into quantitative geometric comparisons.  We applied our framework to data collected as participants watched, and later verbally recounted, a television episode while undergoing functional neuroimaging.  We found that all participants' remembered trajectories reflected the high-level \textit{essence} (i.e., large-scale narrative features) of the episode's trajectory, but participants differred markedly in their memories for low-level details (i.e, small-scale details of the trajectory).  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains transform our ongoing experiences when we encode them into episodic memories, and provides a formal geometric framework for characterizing the complex dynamic content of naturalistic experiences.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences and labeled as having been recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience has little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of precise recalls is often a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener, while the addition of highly specific details may add comparatively little to successful conveyance of an experience.

How might one go about formally characterizing the \textit{essence} of an experience, or whether it has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual ``drift" of our ongoing experience is punctuated by sudden changes, or ``shifts"~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) as playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as geometric \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We further introduce two novel metrics for assessing memory quality: (1) the \textit{precision} with which a participant recounts each event and (2) the \textit{distinctiveness} of each recall event (relative to other recalled events).  We examine how these metrics relate to overall memory performance, and discuss the ways in which they improve upon classic ``proportion-recalled'' measures for analyzing naturalistic memory.  Last, we utilize our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode, and how participants recount it.

\section*{Results}
To characterize the dynamic content of the \textit{Sherlock} episode and participants' subsequent recountings, we used a topic model~\citep{BleiEtal03} to discover the latent themes.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a \textit{topics matrix} whose rows are \textit{topics} (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes what mixture of discovered topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) time segments spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping, sliding windows of (up to) 50 scenes, and treated each window as a single ``document" for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) of topics tended to be ``active" in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events" into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each video event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we parsed the time-varying content of the video, to obtain similar estimates for each participant's recall, we treated each overlapping  ``window" of (up to 10) sentences from their transcript as a ``document," and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participants' recalls to events from the original video, despite different participants using widely varying language to describe the events, and that those descriptions often diverged in content and quality from the video annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appear to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events via HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented" events from the original video, and a (potentially differently ordered) sequence of ``recalled" events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observe patterns comparable to classic effects from the list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). In contrast, we do not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather we see higher memory for specific events distributed approximately evenly throughout the video.

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean clustering score: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean clustering score: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.
% NOTE:
% Semantic and temporal clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.
% - Paxton

>>>>

\author{Andrew C. Heusser\textsuperscript{1, 2, \textdagger}, Paxton C. Fitzpatrick\textsuperscript{1, \textdagger}, and Jeremy R. Manning\textsuperscript{1, *}\\\textsuperscript{1}Department of Psychological and Brain Sciences\\Dartmouth College, Hanover, NH 03755, USA\\\textsuperscript{2}Akili Interactive\\Boston, MA 02110\\\textsuperscript{\textdagger}Denotes equal contribution\\\textsuperscript{*}Corresponding author: Jeremy.R.Manning@dartmouth.edu}

@@

\begin{abstract}
The ways in which our experiences unfold over time are, by nature, incalculably complex.  One effective approach to making sense of this complexity is to characterize experiencess as unique \textit{trajectories} through a certain geometric representational space.  In word embedding spaces, for example, each dimension reflects a concept, such that a given coordinate reflects a weighted mixture of those concepts.  We propose a framework for projecting naturalistic experiences into word embedding spaces, such that the conceptual content of each moment of an experience, and how different moments of the experience relate, are reflected by the \textit{shape} of the experience's trajectory.  By projecting memories of those experiences into the same spaces, one may then geometrically compare the shape of the original experience's trajectory to the shape of how it is remembered later.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the original experience's trajectory.  This translates qualitative questions about how we remember naturalistic experiences into quantitative geometric comparisons.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectory of each participant's recall reflected the high-level \textit{essence} (i.e., large-scale narrative summaries) of the episode's trajectory, but participants differed markedly in their memories for low-level features (i.e, small-scale details).  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains transform ongoing experiences when we encode them into episodic memories, and provides a formal geometric framework for characterizing the complex dynamic content of naturalistic experiences.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experience and singularly labeled as having been recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well controlled, trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience has little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether or not they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of exact recalls is often a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener, while the addition of highly specific details may add comparatively little to successful conveyance of an experience.

How might one go about formally characterizing the \textit{essence} of an experience, and whether it has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual ``drift" of our ongoing experience is punctuated by sudden changes, or ``shifts"~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) as playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here, we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and memories of those experiences) as geometric \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or discarded) in the translation into memory.  We further introduce two novel metrics for assessing memory quality: \textbf{1}) the \textit{precision} with which a participant recounts each event, and \textbf{2}) the \textit{distinctiveness} of each recall event (relative to other recalled events).  We examine how these metrics relate to overall memory performance, and discuss the ways in which they improve upon classic ``proportion-recalled'' measures for analyzing naturalistic memory.  Last, we utilize our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of either the episode or how participants would later recount it.

\section*{Results}
To characterize the dynamic content of the \textit{Sherlock} episode and participants' subsequent recountings, we used a topic model~\citep{BleiEtal03} to discover the episode's latent themes.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a \textit{topics matrix} whose rows are \textit{topics} (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix reflect how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes what mixture of discovered topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening, the location where the scene took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping, sliding windows of (up to) 50 scenes, and treated each window as a single ``document" for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1976 fMRI volumes collected as participants viewed the episode.

@@

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (see Fig.~\topics).  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number (usually one or two) of topics tended to be ``active" in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of ``events" into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods} for additional details).  We then created a stable ``summary'' of the content within each video event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

@@

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogously to how we parsed the time-varying content of the video, to obtain similar estimates for each participant's recall, we treated each overlapping  window of (up to 10) sentences from their transcript as a ``document," and computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participants' recalls to events from the original video, despite different participants using widely varying language to describe the events, and that those descriptions often diverged in content and quality from the video annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail---i.e., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the video correlation matrix, there were substantial off-diagonal correlations.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events via HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented" events from the original video, and a (potentially differently ordered) sequence of ``recalled" events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observed patterns comparable to classic effects from list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). In contrast, we did not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather greater memory for specific events distributed approximately evenly throughout the video.

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.~Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean clustering score: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62,~p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54,~p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean clustering score: 0.787, SEM: 0.018), and that semantic clustering score was also related to both hand-annotated  (Pearson's $r(15) = 0.65,~p = 0.004$) and model-derived (Pearson's $r(15) = 0.63,~p = 0.007$) memory performance.
% NOTE:
% Semantic and temporal clustering are both positively correlated with n_events -- r=0.49; p=0.046 and r=0.61; p=0.009, respectively.
% - Paxton
